Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-13 paper.tex
Some additional fixes following @najmahan's comments

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which are
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform slightly better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.\note {Naj}{what is interesting?}  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}), due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows are equivalent when
integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however, should not
mask \note{Naj}{what do you mean by mask?}the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
task encapsulation and nested workflows could well become
showstoppers \note{Naj}{please don't use these from-film-marketing words?! :-) }depending on the type of targeted
infrastructure. Non-reliable infrastructures \note{Naj}{you mean by the showstoppers, i.e. the nested workflows which can accommodate distributed executions but are limited by scheduling constraints?} , for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in task encapsulation. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor task encapsulation and pool. \note{Naj}{I feel this paragraph is a bit vague, and doesn't distill and contextualize the thorough work described before.}

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases \note{Naj}{Which?}. Also, task encapsulation and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed \note{Naj}{so are you discussing it here, or in the future?}since 
this interaction may not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is problematic because these engines rely
on general-purpose programming languages such as Bash, Octave and
Python, which are much richer than scientific workflow languages.

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations, in particular nested workflows with task
encapsulation.\note{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}

@@

% Impact of the work
To the best of our knowledge, our work is the first of its kind.\note{Naj}{instead of its kind, explain what kind it is, e.g. systematic review of all ...} So
far, the literature on science gateways and workflow engines has
focused on the description of particular systems, or on the
presentation of a particular architecture.  Instead, our analysis
abstracts and evaluates architectural patterns independently from any
particular system. It provides general insight to integrate science
gateways and workflow engines, which we hope will be useful for
software architects.

>>>>

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which are
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service
invocation) perform slightly better when integrating new engines
(\texttt{E$_1$}), which makes them useful architectures for science
gateways that do not focus on a particular engine.\closednote{Naj}{what is
  interesting?}{Tristan}{Rephrased, thanks.}  However, task encapsulation, workflow import (with
service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}), due to the need
to update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows are equivalent when integrating
new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not
surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however,
should not conceal \closednote{Naj}{what do you mean by
  mask?}{Tristan}{I guess 'hide' or 'conceal' are better
  words. Fixed.} the unique characteristics of architectures regarding
this criterion. Nested workflows are the only architecture that can
easily accommodate distributed workflow execution, which can be
critical in some cases. At the same time, the scheduling constraints
created by task encapsulation and nested workflows may become problematic \closednote{Naj}{please don't use these from-film-marketing
  words?! :-) }{Tristan}{OK!} depending on the type of targeted
infrastructure. Non-reliable infrastructures \closednote{Naj}{you mean by
  the showstoppers, i.e. the nested workflows which can accommodate
  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
for example, could hardly cope with workflow engines being wrapped in
computing tasks as done in task encapsulation. The availability of
multiple engine instances could also become a critical feature for
science gateways with important workloads, which would favor task
encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
  vague, and doesn't distill and contextualize the thorough work
  described before.}

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows;
task encapsulation and nested workflows may complexify
fine-grained debugging. \closednote{Rafael}{I would suggest to provide
  some more examples when this could be interesting.}{Tristan}{Done.}
The availability of fine-grained debugging information may also be
critical to efficient user support.

@@

The particular case of interaction \texttt{g} (workflow language
conversion) requires particular attention when implementing a real
system \closednote{Naj}{so are you discussing it here, or in the
  future?}{Tristan}{It's out of scope, a complete paper in itself. Rephrased this part.} since this interaction may not be easily generalized to any
workflow language. For instance, converting FSL, PSOM or Nipype
pipelines to any other workflow language is problematic because these
engines rely on general-purpose programming languages such as Bash,
Octave and Python, which are much richer than scientific workflow
languages.

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations, in particular nested workflows with task
encapsulation.\answerednote{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}{Tristan}{Good idea, will think about it.}

@@

% Impact of the work
To the best of our knowledge, our work is the first to systematically
review and compare software architectures used to integrate workflow
engines in science gateways.\closednote{Naj}{instead of its kind,
  explain what kind it is, e.g. systematic review of all
  ...}{Tristan}{Done.} So far, the literature on science gateways and
workflow engines has focused on the description of particular systems,
or on the presentation of a particular architecture.  Instead, our
analysis abstracts and evaluates architectural patterns independently
from any particular system. It provides general insight to integrate
science gateways and workflow engines, which we hope will be useful
for software architects.
